The performance of presence-based and process-based species distribution models under realistic conditions

### 2020-02-13 ms/ms_latex/ms_sdmComp.tex
discussion paragraphs

% Process-based predictions (lambda, N, etc)
The process-based SDMs generate predictions for a number of biological quantities of interest. For instance, the IPM predicts $\lambda$ in each cell of the landscape. Discrepancies in predicted $\lambda$ values were biased downward in the northern portion of the study region (Fig. \ref{fig:N_Lambda}), more strongly so for the shrub. Mirroring the effect on predicted presence, residual error was most variable and most extensive when \emph{incorrect covariates} were used.

Similarly, the CA models predict the abundance $N$ for each species. Discrepancies in predicted abundance from the CA\textsubscript{p} model showed a spatial banding of upward bias at the range margins for both species (Fig. \ref{fig:N_Lambda}), indicating a tendency to over-predict at range edges. Immediately interior to the range margin, the CA\textsubscript{p} and CA\textsubscript{i} models showed upward bias for the shrub (Fig. \ref{fig:N_Lambda}). The CA\textsubscript{p} model dramatically underestimated abundance of the long-lived shrub with \emph{non-equilibrium} datasets, despite minimal impact on predicted presence or absence. As with the IPM, using \emph{incorrect covariates} tended to result in larger error in abundance predictions for both CA\textsubscript{p} and CA\textsubscript{i}.

@@

% Process-based > Maxent under ideal conditions, and less susceptible to sampling effects
Under ideal conditions, the process-based SDMs, properly matched to the range boundary definition, outperformed Maxent (Fig. \ref{fig:TSSmn}). This generally held true across the sampling scenarios, with minimal impact of \emph{measurement error} or \emph{sampling bias} on the process-based SDMs.

% Climate change and alien species
Critically, there was little or no decrease in performance in the process-based models with \emph{non-equilibrium} samples (Fig. \ref{fig:TSSvIdeal}, B.3, B.4). In fact, the individual-based models (IPM, CA\textsubscript{i}) showed somewhat improved performance for the shrub, as the younger populations had a more even age distribution, allowing for more accurate estimation of the size-dependent vital rate regressions (Fig. B.5a). Because process-based SDMs aim to describe the biological processes that ultimately result in an emergent geographic distribution, care should be taken to select sample populations with an adequately broad range of sizes or ages. In contrast, the performance of Maxent declined the most in the \emph{non-equilibrium} scenario, consistent with past work (CITE). These results thus support the proposal that process-based SDMs are advantageous in situations where populations are not at equilibrium \citep{Merow2011a,Evans2016,Merow2017,Cabral2017}, such as for distributions of alien species or for future distributions as the climate continues to change.

@@

% Dangers of process-based SDMs: dispersal
The accuracy of the CA model predictions were sensitive to dispersal (Fig. \ref{fig:TSSvIdeal}). As expected, underestimating dispersal resulted in smaller predicted ranges, while overestimating dispersal resulted in larger predicted ranges. The predictions for the biennial tended to be a bit more sensitive, particularly in terms of specificity (Fig. B.4). In contrast, the IPM predictions were less affected by errors in the dispersal parameters (Fig. \ref{fig:TSSvIdeal}).

% Other advantages and uses of process-based models
There are a bunch of reasons you might want to build a process-based model aside from increased accuracy. I could draw on some examples from the buckthorn paper.

>>>>

% Process-based predictions (lambda, N, etc)
Discrepancies in predicted $\lambda$ values were biased downward in the northern portion of the study region (Fig. \ref{fig:N_Lambda}), more strongly so for the shrub. Mirroring the effect on predicted presence, residual error was most variable and most extensive when \emph{incorrect covariates} were used. Discrepancies in predicted abundance from the CA\textsubscript{p} model showed a spatial banding of upward bias at the range margins for both species (Fig. \ref{fig:N_Lambda}), indicating a tendency to over-predict at range edges. Immediately interior to the range margin, the CA\textsubscript{p} and CA\textsubscript{i} models showed upward bias for the shrub (Fig. \ref{fig:N_Lambda}). The CA\textsubscript{p} model dramatically underestimated abundance of the long-lived shrub with \emph{non-equilibrium} datasets, despite minimal impact on predicted presence or absence. As with the IPM, using \emph{incorrect covariates} tended to result in larger error in abundance predictions for both CA\textsubscript{p} and CA\textsubscript{i}.

@@

% Process-based > Maxent under ideal conditions, and data scenarios – esp. non-equilibrium
Under ideal conditions, the process-based SDMs, when properly matched to their respective range boundary definitions, outperformed Maxent (Fig. \ref{fig:TSSmn}). The SDMs were also generally unaffected by either the addition of \emph{measurement error} or to geographic \emph{sampling bias}, though the predictions of Maxent were affected somewhat more under both scenarios. Similarly, there was little or no decrease in performance in the process-based SDMs with \emph{non-equilibrium} samples (Fig. \ref{fig:TSSvIdeal}, B.3, B.4). In fact, the individual-based models (IPM, CA\textsubscript{i}) showed marginally improved performance for the shrub, as the younger populations had a more even age distribution, allowing for more accurate estimation of the size-dependent vital rate regressions (Fig. B.5a). Because process-based SDMs aim to describe the biological processes that ultimately result in an emergent geographic distribution, care should be taken to select sample populations with an adequately broad range of sizes or ages. In contrast, the performance of Maxent declined the most in the \emph{non-equilibrium} scenario, consistent with past work (CITE). These results thus support the proposal that process-based SDMs are advantageous in situations where populations are not at equilibrium \citep{Merow2011a,Evans2016,Merow2017,Cabral2017}, such as for distributions of alien species or for future distributions as the climate continues to change.

@@

% Dangers of process-based SDMs: dispersal
The accuracy of the CA model predictions were sensitive to dispersal (Fig. \ref{fig:TSSvIdeal}). As expected, underestimating dispersal resulted in smaller predicted ranges, while overestimating dispersal resulted in larger predicted ranges. The predictions for the biennial tended to be a bit more sensitive, particularly in terms of specificity (Fig. B.4). In contrast, errors in the dispersal parameters had limited impact on the predictions of the IPM (Fig. \ref{fig:TSSvIdeal}). The movement of seeds between cells was insufficient to dramatically extend the boundary of sustainable populations, predicted by $\lambda$ in the IPM. By using both the IPM and CA\textsubscript{i} models, which are constructed using the same environmental and size relationships, sink populations in marginal cells can be distinguished, where individuals exist, but populations are unable to persist.

% Other advantages and uses of process-based models
Apart from potentially increased accuracy, process-based SDMs have other important benefits. While occurrence-based models often predict relative habitat suitability, IPMs predict the growth rate $\lambda$ and CA models predict relative abundance. However, the biological quantities provided by a process-based SDM are much more numerous. For instance, the models used here each predict survival rates, flowering rates, seed production, and germination rates in each cell, either at the level of the individual or as population averages. Further, process-based SDMs are naturally able to incorporate management actions \citep{Szewczyk2019} or a dynamic environment (CITE). Uncertainty in the covariates, for example in the form of climate change scenarios, and uncertainty in the parameters can be incorporated through multiple iterations of the models to better describe the range of plausible outcomes. The flexible nature of process-based SDMs thus provides an important opportunity to evaluate potential actions and plan for an inherently uncertain future.
